---
title: "Approaching the practicum"
author: Romain Ferrali & Paul Love
execute:
    echo: false
    message: false
    warning: false
---

```{r}
#| label: setup
library(tidyverse)
library(haven)
library(modelsummary)

df <- bind_rows(
  read_dta("./data/practicum-jan-2014.dta") |> mutate(after = 0),
  read_dta("./data/practicum-may-2014.dta") |> mutate(after = 1)
) |>
  mutate(
    approval = putinapprov,
    int_yes = 1 - int_nono
  )
```

This document has two parts. The first is general advice on how to approach an open-ended data analysis like the practicum: how to structure the report, how to write for a non-technical audience, how to translate hypotheses into regressions, how to think about figures, and how to be honest about what your data can and cannot establish. The second is a worked example using the practicum data on the Crimea annexation, showing what these recommendations look like in practice. The example is partial — it only chases a handful of the hypotheses you might explore — but it should give you a clear sense of the writing style we are looking for.

# How to approach the analysis

## Tell a story with data

The hardest part of the practicum is not the statistics. It is the writing — and specifically, writing in a way that a non-statistical reader will follow and find interesting. A useful benchmark to keep in mind is your mother (or anyone close to you who does not have a quantitative background). She is intelligent and curious, but she does not know what a regression coefficient is, she does not care about R-squared values, and she will close the document at the first paragraph that bores her. Your job is to tell her a story using data.

A story has a fact to be explained, candidate explanations, evidence weighed against those explanations, and a conclusion that is honest about what we have learned and what we have not. That is the arc your report should follow. Tables and regressions are tools that serve the story; they are not the story.

A second benchmark: dense and short beats long and fluffy. The minimum length for the practicum is about three pages, but you should not pad. If you can make the case in three pages, do not stretch it to five. Every sentence should earn its place.

## Structure the report

A useful structure for the body of the report:

- **Open with the fact you are trying to explain, and quantify it.** "Putin's approval rose after the Crimea annexation" is fine; "Putin gained about 10 percentage points in a couple of months" is much better. If you have the numbers, use them in the very first paragraph. Mentioning the data source in the same sentence buys you credibility for free.

- **Introduce a hypothesis with a transition paragraph.** State the hypothesis. Acknowledge that the available measures are imperfect (rarely does a survey ask the exact question you wish it had asked). Explain how you are operationalizing the idea with the variables you actually have. This single paragraph turns a popular claim into something testable, and it does so without slowing the reader down.

- **Use an "at first sight... but on closer look..." arc when results require nuance.** Start with the visible pattern in the raw data — what the eye picks up. Then explain why that pattern alone is not the whole story (because the difference is not statistically significant, because controls erase it, because a different proxy gives a different answer). End with what we should conclude given everything together.

- **Push technical detail to an appendix.** Variable names from the codebook, the full regression tables, the boring specifications — none of this belongs in the main text. The main text should describe the patterns in plain language. The appendix is where the technically inclined reader can look up the specifications.

## Translate hypotheses into regressions

A useful trick: each hypothesis implies a comparison, and each comparison maps cleanly to a regression.

The hypothesis "people who consume more news rallied to Putin more after the annexation" is not the same as "people who consume more news approve of Putin more." The first implies a particular comparison: the *change* in approval after annexation should be larger among heavier news consumers than among lighter news consumers. The natural regression is `approval ~ after * news_consumption`, where the interaction term captures exactly that differential change. A simple comparison of approval levels by news consumption (without the interaction) would not test the rally hypothesis at all — it would just describe a baseline correlation.

Before you start coding, write down each hypothesis you plan to test, sketch the comparison it implies, and pick the regression specification accordingly. If two hypotheses imply the same regression, consider whether you can really tell them apart from the data.

## Interpret regressions for a non-technical reader

Try to write the body of the report without ever using the word "regression." It is harder than it sounds, and the discipline is good for you. When you would otherwise say "we ran a regression with controls," say something like:

> We compared individuals before and after the annexation who not only had the same level of news consumption, but also the same age, gender, and education level. This makes for more pointed comparisons and improves the precision of our estimates. With these adjustments, we still find...

Two things happen at once: you have explained what controls are, and you have motivated *why* you are adding them, all without invoking jargon. The technical reader knows what you mean. The non-technical reader has learned something.

The same trick works for confidence intervals ("the difference, while large in size, is small enough that it could have arisen by chance"), for fixed effects ("we compared each respondent to themselves over time"), and for interactions ("we asked whether the change in approval after annexation was larger for some groups than for others"). Make a habit of translating before you write.

## Keep controls consistent across specifications

When you present several regressions side by side — typically a base specification, then the same specification with controls — keep the *same* set of controls in every column where controls appear. Do not add age in one column, age and gender in another, and age and education in a third. Cherry-picking control sets is a way to make a coefficient look significant; it is also a tell that something is fishy. If you have decided that age, gender, and education are reasonable controls, use exactly those three throughout.

The same logic applies across hypotheses. If you control for age, gender, and education when testing the news-watching hypothesis, use the same three controls when you test the resentment hypothesis. Consistency makes the comparison across specifications meaningful.

## Make your figures earn their place

Figures should make one point each, and they should make it visible to a reader who has not read your text yet. A few practical rules drawn from the way professional figures are built:

- **Show the raw data, not just the trend line.** Your reader understands points and averages. They do not necessarily understand what a fitted line means. If you only show the line, you have hidden the evidence behind it. If you must show only one thing, show the points; if you can show both, do.

- **Add confidence intervals when comparing means.** They tell the reader, at a glance, whether the difference she sees is real or could plausibly have arisen by chance. Without them, it is on her to guess.

- **Axis breaks should respect the data.** If the variable on the x-axis is "number of news events recalled," it takes integer values. Tick marks at 2.5 do not make sense. Use breaks at 0, 2, 4, ... so the eye lands on points that exist.

- **Y-axis range should let the reader verify your claims.** If your text says "approval hovered around 90%," there had better be a tick mark at 90% so the reader can check. If your text says approval is between 0 and 100%, set the limits to 0 and 100% — do not let ggplot crop in to make the difference look bigger than it is.

- **Captions are long and descriptive.** A figure caption is the figure's title; treat it that way. Spell out what the points represent, what the lines represent, and what the reader should look for. Distinguish raw data from fitted lines explicitly.

- **Use a legend instead of an axis when it is cleaner.** "Before annexation" and "After annexation" do not need to be the x-axis labels of two separate panels. A single panel with a color legend at the bottom often reads better.

## Be honest about causality

A standing reflex when reading a report is to interpret what you see as causal. We almost cannot help it. As the author, your job is to make explicit when a finding *is* causal, when it is merely a description of what happened, and when it is somewhere in between.

For every claim, name the treatment and the outcome. In the practicum, the natural treatment is "the annexation" and the outcome is "approval of Putin." Then ask: do we have a control group? In the practicum, the answer is no — everyone in Russia was exposed to the same news cycle. The headline finding ("approval rose after annexation") is therefore a description of what happened in Russia between January and May 2014. It is not, strictly speaking, the causal effect of the annexation, because we cannot rule out that some other event in those months also moved approval.

When you compare *subgroups* (heavy news consumers vs. light ones, people who dislike the West vs. people who do not), you are doing what is sometimes called subgroup analysis or heterogeneous treatment effects analysis. This can be very informative — but it does not rescue causality. It just tells you whether some kinds of people moved more than others.

There is one more subtle trap worth flagging: the difference between a moderator and a mediator.

- A **moderator** is a stable, pre-existing trait of the respondent that you think shapes how they respond to treatment. A respondent's vote in the 2012 presidential election is a moderator: they cast it years before the annexation, and a question asked in 2014 is just a retrospective measure of something that already happened.
- A **mediator** is something the treatment itself changes, which then affects the outcome. If the war made some respondents angrier at the West, and that anger is what boosted Putin's approval, then "anger at the West" is a mediator on the causal path from treatment to outcome.

The trouble is that a variable measured *after* treatment can easily be a mediator dressed up as a moderator. If you measure "disapproval of the United States" in May 2014 — after the annexation — you cannot tell whether disapproval predates the war (and shapes the response to it) or whether the war boosted disapproval (and disapproval boosted approval of Putin). In the practicum data, only a couple of variables (such as the 2012 vote) are unambiguously pre-treatment; most of the candidate moderators were measured after annexation and could be either moderators or mediators. The cleanest fix would be panel data — surveying the same respondents before and after the annexation — but the practicum does not provide that. Acknowledge the limitation explicitly.

## Acknowledge the limitations

A short paragraph near the end of the report should name the limitations of your analysis. Two recurring kinds:

- **Measurement error.** Almost every variable in a survey is a proxy for something we actually want to measure. The practicum has no direct question about TV watching, even though the propaganda hypothesis is fundamentally about TV. We use internet usage and political-news recall as imperfect substitutes — but internet propaganda is not the same as television propaganda, and being able to recall recent news events is not the same as consuming them. State the gap honestly.

- **Missing data.** Respondents skip questions, and the people who skip are often systematically different from the people who answer. Recoding missing values explicitly (rather than silently dropping them) lets you check whether non-respondents look different from respondents on the variables you do observe. The practicum data has missingness in several places; it is worth checking.

Acknowledging these things does not weaken your report; it strengthens it. A reader who feels you are being straight with them will trust the rest of what you say.

# A worked example

What follows is a partial response to the practicum, written in the voice we are recommending you use. It chases only one of the hypotheses suggested in the assignment — the idea that heavier news consumers rallied more to Putin — but it should give you a feel for how the pieces fit together.

---

Following the annexation of Crimea in February 2014, Vladimir Putin's approval ratings soared. The Levada Center, a respected and reliable polling institution in Russia, conducts regular public opinion polls on large (about 1,600 respondents), representative samples of the Russian population. In their January 2014 survey, Putin's approval rating stood at `r round(mean(df$approval[df$after == 0], na.rm = TRUE) * 100, 1)`%; in their May 2014 survey, it had risen to `r round(mean(df$approval[df$after == 1], na.rm = TRUE) * 100, 1)`%, an increase of roughly ten percentage points in only four months. @fig-overall makes the jump visible.

```{r}
#| label: fig-overall
#| fig-cap: "Putin's approval rating before and after the Crimea annexation. Points show the average rating in each Levada Center wave; vertical bars show 95% confidence intervals."
lm(approval ~ after, data = df) |>
  broom::augment(
    newdata = tibble(after = c(0, 1)),
    interval = "confidence"
  ) |>
  ggplot(aes(x = factor(after), y = .fitted, ymin = .lower, ymax = .upper)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +
  scale_x_discrete(labels = c("Before annexation", "After annexation")) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(x = NULL, y = "Approval rating") +
  theme_minimal()
```

One popular explanation for why the annexation lifted Putin's standing attributes the rise to the Russian state propaganda machine, particularly television: the constant favorable coverage of the annexation supposedly converted neutral Russians into supporters. We test that hypothesis here.

## Was the rally driven by exposure to news?

The Levada surveys do not measure television consumption directly, but several survey items allow us to evaluate the propaganda claim, albeit imperfectly. In particular, the survey asks respondents how many recent news events they can recall, which we use as a proxy for engagement with political news. If the propaganda machine was responsible for the rise in approval, we would expect those who consume more news to show larger increases in approval after the annexation.

@fig-news-consumption shows that after the war, respondents had a level of approval hovering around 90%, regardless of how many news events they recalled. Before the war, by contrast, there was a clear negative relationship between news recall and approval: people who remembered more news events were more likely to disapprove of Putin. This implies that the people most exposed to news saw the largest jump in approval after annexation. At first sight, the pattern supports the propaganda hypothesis: those most exposed to political coverage rallied the most.

```{r}
#| label: fig-news-consumption
#| fig-cap: "Putin's approval before and after the Crimea annexation, by the number of recent news events the respondent could recall. Points and thin lines show raw averages; thick lines show linear trends within each wave."
df |>
  group_by(after, pol_news_count) |>
  summarize(approval = mean(approval, na.rm = TRUE)) |>
  mutate(after = ifelse(after == 0, "Before annexation", "After annexation")) |>
  ggplot(aes(x = pol_news_count, y = approval, color = after)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)
  ) +
  scale_x_continuous(breaks = seq(0, 14, 2)) +
  labs(
    x = "Number of political news events recalled",
    y = "Approval rating",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The insight does not withstand further scrutiny. While the differential increase visible in @fig-news-consumption looks substantial, it is not statistically significant — that is, it is small enough that we cannot rule out chance as the explanation. To make a more pointed comparison, we compared individuals before and after the annexation who shared not only the same level of news recall but also the same gender, age, and education level. Even with these adjustments, the differential increase remains statistically indistinguishable from zero. We checked several other proxies for news exposure — whether the respondent recalled any news event at all, whether they used the internet, and whether they used the internet for news — and reached the same conclusion in each case. None of these comparisons offers convincing evidence that heavy news consumers rallied more than light news consumers (see @tbl-news in the appendix).

## Appendix

```{r}
#| label: tbl-news
#| tbl-cap: "Linear regressions of Putin approval on the interaction between an after-annexation indicator and various proxies for news consumption. Controls include age, gender, and education level. The 'Yes' rows correspond, respectively, to recalling at least one news event, recalling more news events (count), using the internet, and using the internet for news. None of the after-by-news interactions is statistically significant at conventional levels."
list(
  lm(approval ~ after * pol_news, data = df),
  lm(approval ~ after * pol_news + age + female + educ, data = df),
  lm(approval ~ after * pol_news_count, data = df),
  lm(approval ~ after * pol_news_count + age + female + educ, data = df),
  lm(approval ~ after * int_yes, data = df),
  lm(approval ~ after * int_yes + age + female + educ, data = df),
  lm(approval ~ after * int_news, data = df),
  lm(approval ~ after * int_news + age + female + educ, data = df)
) |>
  modelsummary(
    stars = TRUE,
    coef_map = c(
      "after" = "After",
      "pol_news" = "Yes",
      "pol_news_count" = "Yes",
      "int_yes" = "Yes",
      "int_news" = "Yes",
      "after:pol_news" = "After x Yes",
      "after:pol_news_count" = "After x Yes",
      "after:int_yes" = "After x Yes",
      "after:int_news" = "After x Yes"
    ),
    gof_map = c("nobs", "r.squared"),
    add_rows = matrix(
      c("Controls", rep(c("No", "Yes"), 4)),
      nrow = 1
    ) |>
      as.data.frame()
  ) |>
  tinytable::group_tt(
    j = list(
      "Recalls news (binary)" = 2:3,
      "Recalls news (count)" = 4:5,
      "Uses internet" = 6:7,
      "Uses internet for news" = 8:9
    )
  )
```

