---
title: "Lecture 24: Network analysis"
format: revealjs
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false
# install.packages(c("igraph", "igraphdata", "netrankr"))
library(tidyverse)
library(igraph)
library(igraphdata)
library(netrankr)
```

# Networks as tables

## The recipe, again

- Take unstructured data.
- Turn it into a table.
- Apply the regression and visualization tools you already know.

Today's flavor: **networks**.

## Why political scientists care

- Coalitions in a legislature
- Treaties and alliances between states
- Citations between Supreme Court decisions
- Trade flows
- Retweet graphs during a protest
- Donor relationships among campaign donors

. . .

Once each node has features, the regression toolkit applies unchanged.

# Two ingredients

## A network is just

- **Nodes** (vertices) — the entities
- **Edges** (ties) — the relationships between them

. . .

The flavor of the network determines what you can do with it.

## Three axes of variation

- **Directed vs undirected**
- **Weighted vs unweighted**
- **Single-layer vs multi-layer**

. . .

A guessing game with platforms you already use.

## Directed vs undirected

- Twitter follows — **directed** (one-way is allowed)
- Facebook friends — **undirected** (mutual by construction)
- LinkedIn connections — **undirected**
- Snapchat best friends — **directed**, asymmetric
- Phone calls / email — **directed**

. . .

Question: does the relationship go both ways?

## Weighted vs unweighted

- "Are we friends?" → **unweighted** (yes/no)
- "How many messages this month?" → **weighted**
- Trade volume between countries → **weighted**
- Coauthorship counts → **weighted**

. . .

Question: does the edge carry a strength?

## Single- vs multi-layer

- Facebook friends ≠ messages ≠ likes ≠ tagged-in-photo
- Same nodes, several kinds of ties

. . .

Today: single-layer only.

## Why this matters

- Most tools assume a specific flavor.
- Centrality is defined slightly differently on directed vs undirected graphs.
- Many canonical models assume undirected, unweighted, single-layer.

. . .

Always ask: *what kind of network do I have?*

# Storing networks

## Edge list

```{r}
#| echo: true
edges <- tribble(
  ~from , ~to ,
  "A"   , "B" ,
  "A"   , "C" ,
  "B"   , "C" ,
  "C"   , "D" ,
  "D"   , "E"
)
edges
```

One row per edge. Edge attributes go in extra columns.

## Adjacency matrix

```{r}
#| echo: true
g_toy <- graph_from_data_frame(edges, directed = FALSE)
as_adjacency_matrix(g_toy, sparse = FALSE)
```

$n \times n$. Symmetric for undirected graphs. Mostly zeros.

## Edge lists win at scale

- Real networks are **sparse**.
- Facebook: ~3 billion users, ~few hundred friends each.

. . .

- Adjacency matrix: ~$9 \times 10^{18}$ cells
- Edge list: ~$10^{11}$ rows

. . .

Edge lists are the right way to store a network on disk.

## Reading network data

- `graph_from_data_frame(edges, vertices = nodes)` for an edge list
- `graph_from_adjacency_matrix(A)` for a matrix
- Both return the same `igraph` object

# A running example

## Zachary's karate club

- 34 members of a 1970s American university karate club
- Friendship ties collected by Wayne Zachary
- The club **split into two factions** during data collection
- The standard test case in network science

## Loading it

```{r}
#| echo: true
data(karate)
karate
```

`UNW-` = **U**ndirected, **N**amed, **W**eighted. `Faction` is a node attribute.

# Visualizing networks

## Node coordinates aren't in the data

- An edge list says **who** is connected to **whom**.
- It does *not* say where to put each node on a page.

. . .

Layout algorithms decide.

## Default `igraph` layout

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 5
set.seed(2026)
plot(karate, vertex.size = 12, vertex.label.cex = 0.7)
```

## Force-directed (Fruchterman–Reingold)

- Connected nodes attract.
- Disconnected nodes repel.
- The two factions emerge as visual clusters — *the algorithm did not know about them*.

## Tree layout

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 5
plot(
  karate,
  layout = layout_as_tree(karate),
  vertex.size = 12,
  vertex.label.cex = 0.7
)
```

## Same data, different story

- The karate club is **not** a tree.
- The layout imposed a hierarchy that isn't there.
- Right call for trees; wrong for almost anything else.

. . .

Layout = an authorial choice.

## Polish: degree-sized, faction-colored

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 5
set.seed(2026)
plot(
  karate,
  vertex.size = 5 + 1.5 * degree(karate),
  vertex.color = ifelse(V(karate)$Faction == 1, "steelblue", "tomato"),
  vertex.label.cex = 0.7,
  edge.width = 0.5
)
```

# Networks in regression

## From network to regressors

- Extract a **node table**: one row per node.
- Each derived feature is a column.
- The rest is regression as usual.

## Three centrality measures

- **Degree** — number of edges. Popularity / activity.
- **Betweenness** — how often you sit on shortest paths. Broker / gatekeeper.
- **Closeness** — average graph-distance to everyone. Information-speed.

## Computing them

```{r}
#| echo: true
karate_nodes <- tibble(
  member = V(karate)$name,
  faction = V(karate)$Faction,
  degree = degree(karate),
  between = betweenness(karate),
  closeness = closeness(karate)
)

karate_nodes |> arrange(desc(degree)) |> head(8)
```

Mr Hi (instructor) and John A (president) lead — the two factional leaders.

# Peer effects

## The question

> Does my outcome depend on my **neighbors'** outcomes?

Examples:

- Friends start running marathons → do you?
- Trading partners adopt the euro → do you?
- A close friend becomes obese → do you?

## Linear-in-means model

$$
y_i = \alpha + \beta x_i + \gamma\, \bar y_{N(i)} + \varepsilon_i
$$

- $\bar y_{N(i)}$ — average outcome among $i$'s network neighbors.
- $\gamma$ — peer-effect coefficient.

. . .

Same shape as the inverse-distance regression from lecture 23 — only the meaning of "neighbor" changed.

## A real network with a real outcome

**Padgett's Florentine families (1430s):**

- 16 elite families in Renaissance Florence
- Marriage network from genealogical records
- Wealth in 1000s of lira from the 1427 *catasto* (tax records)
- The data behind Padgett & Ansell's classic *"Robust action and the rise of the Medici"* (1993)

## Loading the data

The dataset ships with R's `network` package — the canonical home of social-network data on CRAN.

```{r}
#| echo: true
#| message: false
data(florentine_m, package = "netrankr")
flo <- florentine_m
flo
```

`wealth`, `priorates`, `totalties` are real attributes from the 1427 *catasto*, attached to the network by the package authors.

## The marriage network

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 5
set.seed(2026)
plot(
  flo,
  vertex.size = 4 + sqrt(V(flo)$wealth) * 1.5,
  vertex.label.cex = 0.7,
  edge.width = 0.6
)
```

Pucci (wealth 3) is isolated — no marriage ties.

## Building the "neighbors' mean" regressor

```{r}
#| echo: true
A <- as_adjacency_matrix(flo, sparse = FALSE)
deg <- pmax(rowSums(A), 1) # avoid divide-by-zero for Pucci
W <- A / deg # row-normalized adjacency

flo_data <- tibble(
  family = V(flo)$name,
  wealth = V(flo)$wealth,
  neighbor_wealth = as.numeric(W %*% V(flo)$wealth),
  has_neighbors = rowSums(A) > 0
)

head(flo_data)
```

One matrix multiplication = the neighbor-mean column for the entire dataset.

## The regression

```{r}
#| echo: true
flo_reg <- flo_data |> filter(has_neighbors)

lm(wealth ~ neighbor_wealth, data = flo_reg) |>
  summary()
```

A positive, large coefficient on `neighbor_wealth`.

## What did we just estimate?

- Families connected to wealthy in-laws are themselves wealthy.
- $\hat\gamma > 0$, statistically significant.

. . .

But did wealth **spread through marriage**, or did wealthy families **marry each other**?

# Why peer effects are hard

## The reflection problem (Manski 1993)

- Your friends affect you.
- You affect your friends.
- $y_i$ and $\bar y_{N(i)}$ are determined together.

. . .

A regression coefficient comes out — its causal interpretation is muddled.

## Homophily

- People connect to others **like them**.
- Rich families marry rich families.
- The correlation reflects **selection**, not influence.

## A devastating gut-check

**Cohen-Cole & Fletcher (2008)** ran the Christakis–Fowler peer-effects regression on:

- acne
- height
- headaches

. . .

They found significant "peer effects" in all three.

. . .

None of these can spread through friendships.

## What actually identifies peer effects

A design that breaks selection:

- Random roommate assignment in dorms
- Randomized seeding in a development experiment
- Exogenous shocks that hit some neighbors but not others

. . .

Without one of those, treat $\hat\gamma$ as **descriptive**, not causal.

# Wrap-up

## Take home

- Networks = a node table with derived features.
- New regressors: **centrality** (degree, betweenness, closeness) and **neighbors' means**.
- Layouts are an authorial choice; force-directed is the safe default.
- Peer-effects coefficients are easy to estimate and hard to interpret.

## Skipped on purpose

- Directed-network methods
- Dynamic networks (ties forming and dissolving)
- Community detection
- Network-generating models (Erdős–Rényi, preferential attachment, ERGMs)

`igraph` will be there when you need them.
