---
title: "Lecture 20-21: Time series"
format: revealjs
---

```{r}
#| label: setup
library(tidyverse)
library(haven)
library(modelsummary)
library(fixest)

df <- bind_rows(
  read_dta("./data/practicum-jan-2014.dta") |> mutate(after = 0),
  read_dta("./data/practicum-may-2014.dta") |> mutate(after = 1)
) |>
  # data cleaning
  mutate(
    approval = putinapprov,
    int_yes = 1 - int_nono,
    us_disapprov = 1 - usapprov,
    eu_disapprov = 1 - euapprov,
    ukraine_disapprov = 1 - ukrapprov
  )
```

Time series data is data with a time element. We also call this dynamic data (vs. static/cross-section data). Can be: 

- Panel data (units you follow over time; e.g., survey the same people every year, or track the same countries' GDP every year)
- Repeated cross-section data (multiple units observed at multiple time points, but not the same units each time; e.g., practicum data)


## Uses of time-series data

- Forecasting (in class: find examples of things useful to forecast)
- Improve causal inference: 
  - Understand the importance of events (e.g. how did the 2008 financial crisis affect the economy?)
  - Understand treatment effect dynamics (e.g. how long does a policy's effect last?)
  - Estimate causal effects even when treatment is not randomly assigned

## Analyzing repeated cross-section data: the practicum case

Eight years before the Russian full-scale invasion into Ukraine, on February 27th of 2014, Vladimir Putin illegally annexed the Ukrainian peninsula of Crimea. His approval ratings soared. 

Why did the annexation of Crimea boost his approval? 

- One argument is that the Russian state propaganda machine, mostly through television, was an important factor. 
- Another argument is that resentment toward the West, and in particular Western sanctions against Russia, led people to view Crimea's annexation as a blow against their oppressors. 
- Yet others argue that it was mere territorial expansion that made people approve of Putin. 

## Battle plan

- One argument is that the Russian state propaganda machine, mostly through television, was an important factor. 
  - We can test this by looking at whether people who watch more news increase their approval more after annexation.
- Another argument is that resentment toward the West, and in particular Western sanctions against Russia, led people to view Crimea's annexation as a blow against their oppressors. 
  - We can test this by looking at whether people who are more resentful of the West increase their approval more after annexation.
- Yet others argue that it was mere territorial expansion that made people approve of Putin. 
  - We can test this by looking at whether people who are more nationalistic increase their approval more after annexation.

# Establishing the baseline: did approval increase?

# Simple visualization

```{r}
#| label: approval-plot simple
df |>
  group_by(after) |>
  summarize(approval = mean(approval, na.rm = T)) |>
  ggplot(aes(x = after, y = approval)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = c(0, 1), labels = c("Before", "After")) +
  labs(
    x = "Time",
    y = "Mean approval",
    title = "Putin's approval before and after Crimea annexation"
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  theme_minimal()
```

## With regression

$y_{it}$, Putin's approval rating for individual $i$ at time $t \in \{\text{before} = 0, \text{after} = 1\}$. 

$$
y_{it} = \alpha + \beta \cdot t + \epsilon_{it}
$$

. . .

- $\alpha$ is the mean approval before annexation (when $t=0$).
- $\alpha + \beta$ is the mean approval after annexation (when $t=1$).
- $\beta$ is the change in approval after annexation (when $t=1$). This is the key parameter of interest: if $\beta > 0$, then approval increased after annexation.

```{r}
#| label: approval-reg simple
lm(approval ~ after, data = df) |>
  modelsummary(
    stars = TRUE,
    coef_rename = c("after" = "After annexation"),
    gof_map = c("nobs", "r.squared")
  )
```

## Visualizing the regression results

```{r}
#| label: approval-plot
lm(approval ~ after, data = df) |>
  broom::augment(
    newdata = tibble(after = c(0, 1)),
    interval = "confidence"
  ) |>
  ggplot(aes(x = after, y = .fitted, ymin = .lower, ymax = .upper)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(), width = 0.1) +
  scale_x_continuous(breaks = c(0, 1), labels = c("Before", "After")) +
  labs(
    x = "Time",
    y = "Mean approval",
    title = "Putin's approval before and after Crimea annexation"
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  theme_minimal()
```

# Testing the resentment hypothesis

## Candidate variables

- `us_disapprov`: disapproval of the United States.
- `eu_disapprov`: disapproval of the European Union.
- `dsxnshrm2`: belief that Western sanctions harm ordinary people in Russia. -> only asked in one wave, so we can't use it.

## Simple visualization

```{r}
#| label: approval-plot resentment
df |>
  group_by(after, us_disapprov) |>
  summarize(approval = mean(approval, na.rm = T)) |>
  mutate(us_disapprov = ifelse(us_disapprov == 1, "Yes", "No")) |>
  ggplot(aes(
    x = after,
    y = approval,
    color = us_disapprov,
    group = us_disapprov
  )) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = c(0, 1), labels = c("Before", "After")) +
  labs(
    x = "Time",
    y = "Mean approval",
    title = "Putin's approval before and after Crimea annexation"
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  theme_minimal()
```

## With regression

$y_{it}$, Putin's approval rating for individual $i$ at time $t \in \{\text{before} = 0, \text{after} = 1\}$.

$x_{it}$, whether individual $i$ disapproves of the United States at time $t$.

$$
y_{it} = \alpha + \beta_1 t + \beta_2 x_{it} + \beta_3 t x_{it} + \epsilon_{it}
$$

. . .

- $\alpha$: approves of US, before.
- $\alpha + \beta_1$: approves of US, after.
- $\alpha + \beta_2$: disapproves of US, before.
- $\alpha + \beta_1 + \beta_2 + \beta_3$: disapproves of US, after.
- $\beta_1$: change in approval after annexation for those who approve of the US.
- $\beta_1 + \beta_3$: change in approval after annexation for those who disapprove of the US.
- $\beta_3$: extra change in approval after annexation for those who disapprove of the US.
- If $\beta_3 > 0$, resentful respondents rallied more around Putin than non-resentful respondents after annexation.

. . .

```{r}
#| label: approval-reg resentment
lm(approval ~ after * us_disapprov, data = df) |>
  modelsummary(
    stars = TRUE,
    coef_rename = c(
      "after" = "After annexation",
      "us_disapprov" = "Disapproves of US"
    ),
    gof_map = c("nobs", "r.squared")
  )
```

## Robustness (1): adding controls

```{r}
#| label: approval-reg resentment controls
list(
  lm(approval ~ after, data = df),
  lm(approval ~ after * us_disapprov, data = df),
  lm(approval ~ after * us_disapprov + age + female + educ, data = df)
  # add regression with us_disapprov having explicitly coded NAs + controls
) |>
  modelsummary(
    stars = TRUE,
    coef_rename = c(
      "after" = "After annexation",
      "us_disapprov" = "Disapproves of US",
      "age" = "Age",
      "female" = "Female",
      "educ" = "Education"
    ),
    gof_map = c("nobs", "r.squared")
  )
```

## Robustness (2): scaling up

```{r}
#| label: robustness resentment
list(
  lm(approval ~ after * us_disapprov, data = df),
  lm(approval ~ after * us_disapprov + age + female + educ, data = df),
  lm(approval ~ after * eu_disapprov, data = df),
  lm(approval ~ after * eu_disapprov + age + female + educ, data = df)
) |>
  modelsummary(
    stars = TRUE,
    coef_map = c(
      "after" = "After",
      "us_disapprov" = "Yes",
      "eu_disapprov" = "Yes",
      "after:us_disapprov" = "After x Yes",
      "after:eu_disapprov" = "After x Yes"
    ),
    gof_map = c("nobs", "r.squared"),
    add_rows = matrix(
      c("Controls", rep(c("No", "Yes"), 2)),
      nrow = 1
    ) |>
      as.data.frame()
  ) |>
  tinytable::group_tt(
    j = list(
      "Disapproves of US" = 2:3,
      "Disapproves of EU" = 4:5
    )
  )
```


# Other hypotheses

## Candidate variables


- News watching:
  - `int_yes`: has internet (best proxy for news-watching that we have).
  - `int_news`: watches news on the internet.
  - `pol_news`: remembers political events in the news.
  - `pol_news_count`: number of political events in the news that the respondent remembers.
- Nationalism:
  - `hyp_ldpr`: would vote for the LDPR (the most nationalist party in Russia).
  - `vote_zhirin`: voted for Zhirinovsky in 2012 (the most nationalist candidate in the 2012 presidential election).

## News watching hypothesis

```{r}
#| label: robustness
list(
  lm(approval ~ after * int_yes, data = df),
  lm(approval ~ after * int_yes + age + female + educ, data = df),
  lm(approval ~ after * int_news, data = df),
  lm(approval ~ after * int_news + age + female + educ, data = df),
  lm(approval ~ after * pol_news, data = df),
  lm(approval ~ after * pol_news_count, data = df),
  lm(approval ~ after * pol_news + age + female + educ, data = df),
  lm(approval ~ after * pol_news_count + age + female + educ, data = df)
) |>
  modelsummary(
    stars = TRUE,
    coef_map = c(
      "after" = "After",
      "int_yes" = "Yes",
      "int_news" = "Yes",
      "pol_news" = "Yes",
      "pol_news_count" = "Yes",
      "after:int_yes" = "After x Yes",
      "after:int_news" = "After x Yes",
      "after:pol_news" = "After x Yes",
      "after:pol_news_count" = "After x Yes"
    ),
    gof_map = c("nobs", "r.squared"),
    add_rows = matrix(
      c("Controls", rep(c("No", "Yes"), 4)),
      nrow = 1
    ) |>
      as.data.frame()
  ) |>
  tinytable::group_tt(
    j = list(
      "Internet" = 2:3,
      "Internet for news" = 4:5,
      "Remembers pol. evts" = 6:7,
      "Remembers (count)" = 8:9
    )
  )
```

## Nationalism hypothesis

```{r}
#| label: robustness territorial
list(
  lm(approval ~ after * ukraine_disapprov, data = df),
  lm(approval ~ after * ukraine_disapprov + age + female + educ, data = df),
  lm(approval ~ after * hyp_ldpr, data = df),
  lm(approval ~ after * hyp_ldpr + age + female + educ, data = df),
  lm(approval ~ after * vote_zhirin, data = df),
  lm(approval ~ after * vote_zhirin + age + female + educ, data = df)
) |>
  modelsummary(
    stars = TRUE,
    coef_map = c(
      "after" = "After",
      "ukraine_disapprov" = "Yes",
      "hyp_ldpr" = "Yes",
      "vote_zhirin" = "Yes",
      "after:ukraine_disapprov" = "After x Yes",
      "after:hyp_ldpr" = "After x Yes",
      "after:vote_zhirin" = "After x Yes"
    ),
    gof_map = c("nobs", "r.squared"),
    add_rows = matrix(
      c("Controls", rep(c("No", "Yes"), 3)),
      nrow = 1
    ) |>
      as.data.frame()
  ) |>
  tinytable::group_tt(
    j = list(
      "Disapproves of Ukraine" = 2:3,
      "Would vote LDPR" = 4:5,
      "Voted Zhirinovsky 2012" = 6:7
    )
  )
```

# Overall results

- No evidence for the news-watching hypothesis.
- Mixed evidence for the resentment hypothesis: disapproval of the US seems to be associated with a bigger increase in approval after annexation, but disapproval of the EU does not.
- Stronger evidence for the nationalism hypothesis: support for the ultranationalist party associated with a bigger increase in approval after annexation.

# Limitations

- Measurement error: our measures of news-watching is imperfect (internet instead of TV)
- Missing data: we should have checked looked into missingness (e.g. use explicity NAs).
- Causality

# Causality

## Our treatment: annexation of Crimea.

- What is our control group? None: we don't have a group of people who were not exposed to the annexation. We only have before and after measures for the same population.
- So we can't establish causality: we can't rule out that some other event that happened around the same time as the annexation caused the increase in approval.

## Moderators: resentment, nationalism, news-watching.

- We measure most of our moderators at the time of the survey, so for the "after" survey, we measure it after the annexation. This means that we can't rule out that the annexation caused an increase in resentment or nationalism, which in turn caused an increase in approval. In other words, we can't rule out that resentment and nationalism are **mediators** rather than **moderators** of the effect of annexation on approval.
- Only one of our moderators has been measured before annexation (vote for Zhirinovsky)
- $\Rightarrow$ **panel** data with the same respondents before and after annexation would allow us to measure all moderators before annexation and thus establish causality more convincingly.

# Establishing causality with panel data

## A new running example

We leave Crimea and move to Denmark in 2010–2011.

- In 2011 the Danish **Liberal Party** (*Venstre*, "V") suddenly reversed its long-standing position on two welfare policies:
  - halving the maximum unemployment benefits period from 4 to 2 years,
  - abolishing the early-retirement pension (*efterløn*).
- The **Danish People's Party** (*Dansk Folkeparti*, "DF") supported the Liberals on unemployment benefits.
- A five-wave panel survey happened to be in the field *right around* these switches (Slothuus & Bisgaard 2021, AJPS).

. . .

Question for us: **did Liberal-leaning voters change their opinions when their party changed its position?**

## The data

```{r}
#| label: danish-load
danish <- read_csv("data/lecture-20-21-danish.csv", show_col_types = FALSE)
danish
```

- `respondent_id` — 2,902 unique Danes
- `wave` — 1 to 5 (five survey waves in 2010–2011)
- `party` — the respondent's party at wave 1
- `benefit_cut_support` — 0–1 scale, higher = more support for cutting unemployment benefits

## Why this is a better setup than Crimea

- Same people, surveyed repeatedly $\Rightarrow$ we can tell time-varying changes from stable personal characteristics.
- A treatment group: Liberal and Danish People's Party supporters — their party switched.
- A control group: supporters of parties that *didn't* switch.

. . .

If Liberal supporters shift toward shorter benefits between two waves, is that enough to claim the party caused the shift? What else could have happened in the meantime?

## A first look

We focus on waves 2 and 3, bracketing the Liberal Party's switch on unemployment benefits. Let's plot the average support for cutting unemployment benefits, separately for treated (Liberal / DPP) and control (other parties) respondents, before and after the switch.

```{r}
#| label: first-look
danish |>
  filter(wave %in% c(2, 3), !is.na(party)) |>
  mutate(
    group = ifelse(party %in% c("V", "DF"), "Liberal / DPP", "Other parties"),
    wave_label = ifelse(wave == 2, "Before switch", "After switch")
  ) |>
  group_by(group, wave_label) |>
  summarize(
    mean_support = mean(benefit_cut_support, na.rm = TRUE)
  ) |>
  ggplot(aes(
    x = factor(wave_label, levels = c("Before switch", "After switch")),
    y = mean_support,
    color = group,
    group = group
  )) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    x = NULL,
    y = "Support for cutting unemployment benefits",
    color = NULL,
    title = "Did Liberal/DPP supporters move more than others?"
  ) +
  theme_minimal()
```

## Is this causal?

- What does the *gap between the two lines* measure at each wave?
- What does the *change in the gap* from before to after measure?
- What would have to be true for the widening of the gap to be the causal effect of the Liberal Party's position switch?


# Difference-in-differences

## Two bad comparisons

We want to know: **what effect did the Liberal and DPP policy position change have on their supporters' support for said policy?**

. . .

Two bad comparisons: 

1. Treated group (after) - Treated group (before)
2. Treated group (after) - Control group (after)

. . .

1. Bad because other things could have changed between before and after, besides the party switch. Maybe the economy got worse, or there was a news event that made people more supportive of cutting benefits.
2. Bad because the treated and control groups are different people, so their support levels could differ for reasons unrelated to the party switch.

## Combining them makes for a good comparison

We compare:

- the pre-to-post change among **treated** respondents (Liberal + Danish People's supporters)
- to the pre-to-post change among **control** respondents (everyone else)

**Key takeaway.** We use the control group's pre-to-post change to estimate the "secular trend"/counterfactual — how much treated respondents would have moved too, even without the party switch. Then we see if treated respondents moved *more* than that.

## The 2 × 2 table

```{r}
#| label: did-means
danish_23 <- danish |>
  filter(wave %in% c(2, 3), !is.na(party)) |>
  mutate(
    treated = as.integer(party %in% c("V", "DF")),
    post = as.integer(wave == 3)
  )

danish_23 |>
  group_by(treated, post) |>
  summarize(
    mean_support = mean(benefit_cut_support, na.rm = TRUE),
    .groups = "drop"
  )
```

Four numbers — one for each (treated, post) cell.

## The difference-in-differences estimator

Using the four cell means:

$$
\widehat{\text{DiD}} = \bigl(\bar y_{T,\text{post}} - \bar y_{T,\text{pre}}\bigr) - \bigl(\bar y_{C,\text{post}} - \bar y_{C,\text{pre}}\bigr)
$$

- First term: how much treated respondents moved.
- Second term: how much control respondents moved (the "secular trend" — everything that would have happened to treated respondents too, even without the party switch).
- Difference: the *extra* movement among the treated, attributable to the treatment.

## As a regression

The 2 × 2 difference-in-differences is just a regression with an interaction:

```{r}
#| label: did-reg
lm(benefit_cut_support ~ treated * post, data = danish_23) |>
  modelsummary(
    stars = TRUE,
    gof_map = c("nobs", "r.squared")
  )
```

- Intercept: control group, before.
- `treated`: control-vs-treated gap *before* the switch.
- `post`: pre-to-post change for the control group.
- `treated:post`: **the difference-in-differences** — extra movement for treated respondents.

## The assumption we rely on

For the difference-in-differences to be the causal effect of the party switch, we need to believe:

> In the absence of the switch, Liberal/DPP supporters would have moved on this issue by the same amount as supporters of other parties.

. . .

i.e., the time dynamics of the control group are a credible counterfactual for the time dynamics of the treated group, had they not been treated.

. . .

This is called the **parallel trends assumption**.

. . .

We can't test it directly — it's a counterfactual. But we can look at whether the two groups were moving in parallel *before* the treatment.

## A plausibility check, eyeball test

We have an earlier wave (wave 1) — before wave 2. If parallel trends holds, the treated and control groups should have moved roughly in parallel from wave 1 to wave 2, when nothing special was happening.

```{r}
#| label: pretrend-plot
danish |>
  filter(!is.na(party), wave <= 3) |>
  mutate(
    treated = ifelse(party %in% c("V", "DF"), "Treatment", "Control")
  ) |>
  group_by(treated, wave) |>
  summarize(
    mean_support = mean(benefit_cut_support, na.rm = TRUE)
  ) |>
  ggplot(aes(x = wave, y = mean_support, color = factor(treated))) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = 1:3, labels = c("Wave 1", "Wave 2", "Wave 3")) +
  labs(
    x = NULL,
    y = "Support for cutting unemployment benefits",
    color = NULL,
    title = "Pre-treatment trends"
  )
```

## A plausibility check, now with regression

```{r}
#| label: pretrend-reg
danish |>
  filter(wave %in% c(1, 2), !is.na(party)) |>
  mutate(
    treated = as.integer(party %in% c("V", "DF")),
    post12 = as.integer(wave == 2)
  ) |>
  lm(benefit_cut_support ~ treated * post12, data = _) |>
  modelsummary(
    stars = TRUE,
    gof_map = c("nobs", "r.squared")
  )
```

A small, statistically insignificant interaction is reassuring (but not proof). A big one would be a warning sign.

. . .

Classroom discussion: what real-world process could make parallel trends fail here?

# Unit fixed effects

## Using the panel structure

We have a difference-in-differences estimate. But notice: our regression so far doesn't use the fact that we see the *same* respondents at wave 2 and wave 3. We're just treating the data as two independent cross-sections of treated and control respondents.

. . .

Unit fixed effects let us lean on the panel structure directly: we give each respondent their own intercept, which soaks up **everything stable about them** — deep partisanship, personality, childhood, education, …

. . .

That sharpens our design. Let's build it up.

## We already know dummy variables

Back in the regression lectures we added dummies to shift the intercept:

$$
y_i = \alpha + \beta_1 \cdot \text{female}_i + \beta_2 x_i + \varepsilon_i
$$

The dummy `female` adds a constant to women's predicted outcome — essentially, women get their own intercept.

. . .

- Males: $y_i = \underbrace{\alpha}_{\alpha_{\text{male}}} + \beta_2 x_i + \varepsilon_i$
- Females: $y_i = \underbrace{(\alpha + \beta_1)}_{\alpha_{\text{female}}} + \beta_2 x_i + \varepsilon_i$

. . .

**What if we had one dummy per individual?**

## A tiny toy panel

```{r}
#| label: toy-panel
toy <- tribble(
  ~id , ~wave  , ~opinion ,
  "A" , "pre"  ,        2 ,
  "A" , "post" ,        4 ,
  "B" , "pre"  ,        3 ,
  "B" , "post" ,        5 ,
  "C" , "pre"  ,        8 ,
  "C" , "post" ,        9 ,
  "D" , "pre"  ,        7 ,
  "D" , "post" ,        9
) |>
  mutate(
    post = as.integer(wave == "post"),
    id = factor(id)
  )

toy
```

Four people, two waves each. Notice:

- Person C starts much higher than person A — stable personal differences.
- Everyone moves *up* from pre to post — but by different amounts.

## Regression with one dummy per person

```{r}
#| label: toy-fe
lm(opinion ~ id + post, data = toy) |>
  summary()
```

- The coefficients on `idB`, `idC`, `idD` are person-specific intercepts — how much that person's baseline differs from person A's.
- The coefficient on `post` is the average *within-person* change from pre to post: A went up by 2, B by 2, C by 1, D by 2, so the average is 1.75.

. . .

- Person A: $y_{A,t} = \underbrace{\text{Intercept}}_{\alpha_A} + \beta \cdot \text{post}_t + \varepsilon_{A,t}$
- Person B: $y_{B,t} = \underbrace{(\text{Intercept} + \text{idB})}_{\alpha_B} + \beta \cdot \text{post}_t + \varepsilon_{B,t}$

. . .

Those person-specific intercepts are called **unit fixed effects**.

## What unit fixed effects do

A unit fixed effect gives each person their own intercept that soaks up **everything stable about them**:

- personality, childhood, genetics, permanent income, deep partisanship, upbringing, education (for adults), …
- anything we can't measure, too.

. . .

As long as those characteristics don't change over time, they **cannot bias** the coefficient on a variable that *does* change over time — the person's fixed effect has already absorbed them.

. . .

This is exactly what we couldn't do with Crimea: we never saw the same respondent twice.

## The fixed-effects model

- $y_{it}$: $i$'s outcome at time $t$
- $x_{it}$: $i$'s time-varying characteristics at time $t$ (e.g. news consumption this month, current income)
- $x_i$: $i$'s time-invariant characteristics (e.g. year of birth, childhood socialization, deep personality traits)

. . .

$$
y_{it} = \alpha + \beta_1 \text{income}_{it} + \beta_2 \text{female}_i + \beta_3 \text{birthplace}_i + \varepsilon_{it}
$$

. . .

Fixed effects throw away all the $x_i$ variation by giving each person their own intercept:

$$
y_{it} = \alpha_i + \beta_1 \text{income}_{it} + \varepsilon_{it}
$$

. . . 

**Important.** you cannot include any time-invariant variable (e.g. `female`, `birthplace`) in a fixed-effects regression, because their effect is already absorbed by the unit fixed effect. The following regression cannot be estimated:

$$
y_{it} = \alpha_i + \beta_1 \text{income}_{it} + \beta_2 \text{gender}_i + \varepsilon_{it}
$$

## Using fixed effects in difference-in-differences

Back to our difference-in-differences regression. We were working with:

$$
y_{it} = \alpha + \beta_1 \cdot \text{post}_t + \beta_2 \cdot (\text{treated}_i \times \text{post}_t) + \beta_3 \cdot \text{treat}_i + \varepsilon_{it}
$$

This boils down to two regressions:

- Control group: $y_{it} = \underbrace{(\alpha)}_{\alpha_{\text{control}}} + \beta_1 \cdot \text{post}_t + \varepsilon_{it}$
- Treated group: $y_{it} = \underbrace{(\alpha + \beta_2)}_{\alpha_{\text{treatment}}} + (\beta_1 + \beta_3) \cdot \text{post}_t + \varepsilon_{it}$

. . .

Instead, we can give each respondent their own intercept and compare them to themselves over time:

- Control group: $y_{it} = \underbrace{\alpha_i}_{\alpha_{\text{control}, i}} + \beta_1 \cdot \text{post}_t + \varepsilon_{it}$
- Treated group: $y_{it} = \underbrace{(\alpha_i)}_{\alpha_{\text{treatment}, i}} + (\beta_1 + \beta_3) \cdot \text{post}_t + \varepsilon_{it}$

Together:

$$
y_{it} = \alpha_i + \beta_1 \cdot \text{post}_t + \beta_2 \cdot (\text{treated}_i \times \text{post}_t) + \varepsilon_{it}
$$

. . .

Notice what's *missing*: the standalone `treated` term. It's already absorbed by $\alpha_i$, because `treated` doesn't change over time for a given person — it's just one of their stable characteristics. The fixed effect has already done that work.

## The within-group view

Another way to think about it: for each group (here, a person), subtract their own average from their outcome.

```{r}
#| label: toy-within
toy <- toy |>
  group_by(id) |>
  mutate(
    opinion_demeaned = opinion - mean(opinion),
    post_demeaned = post - mean(post)
  ) |>
  ungroup()
toy
```

Running `lm(opinion_demeaned ~ post_demeaned)` on the demeaned data gives *exactly* the same `post` coefficient as the dummy-variable regression. Unit fixed effects = "subtract each person's mean, then regress."

. . .

In words: unit fixed effects throw away all the *between-group* variation and use only the *within-group* movement to estimate the treatment effect. Any stable personal characteristic gets subtracted out in the demeaning step and cannot bias the treatment coefficient.

## Let's double check

```{r}
#| label: toy-demean-fe
list(
  lm(opinion ~ id + post, data = toy),
  lm(opinion_demeaned ~ post_demeaned, data = toy)
) |>
  modelsummary(
    stars = TRUE,
    coef_rename = c("post" = "Post", "post_demeaned" = "Post (demeaned)"),
    gof_map = c("nobs", "r.squared")
  )
```

## Doing this efficiently with `fixest`

With 2,902 Danes, writing out 2,902 dummies is a bad idea. The `fixest` package handles fixed effects efficiently (using the "within" transformation under the hood, so it doesn't actually create 2,902 dummy variables):

```{r}
#| label: fixest-toy
feols(opinion ~ post | id, data = toy)
```

The `| id` part says "absorb a unit fixed effect for `id`." Same estimate as `lm(opinion ~ id + post, data = toy)`, but fast.

## On the real data

Let's run our difference-in-differences two ways: the plain interaction regression, then the same regression with a unit fixed effect per respondent.

```{r}
#| label: fixest-danish
list(
  "DiD" = feols(benefit_cut_support ~ treated * post, data = danish_23),
  "DiD + unit FE" = feols(
    benefit_cut_support ~ post + treated:post | respondent_id,
    data = danish_23
  )
) |>
  modelsummary(
    stars = TRUE,
    coef_map = c(
      "treated:post" = "Treated x Post",
      "post:treated" = "Treated x Post",
      "post" = "Post",
      "treated" = "Treated"
    ),
    gof_map = c("nobs", "r.squared")
  )
```

- Same treatment-effect coefficient. The difference-in-differences number is unchanged.
- But the unit fixed effect has absorbed every stable characteristic of each respondent — we're now identifying the effect from *within-person* movement only.
- And notice the `treated` main effect disappeared in the second column: it's been swallowed by the respondent fixed effects.

# Time fixed effects

## What was `post` actually doing?

Look back at our regression:

$$
y_{it} = \alpha + \beta_1 \cdot \text{treated}_i + \beta_2 \cdot \text{post}_t + \beta_3 \cdot (\text{treated}_i \times \text{post}_t) + \varepsilon_{it}
$$

- `post` is just a dummy for the second wave.
- It's a **time fixed effect** with only two values.

. . .

Time fixed effects absorb anything that hit *everyone* at the same moment — an economic shock, a news cycle, the weather.

## Time fixed effects in general

With more waves, we'd want one dummy per wave:

$$
y_{it} = \alpha_i + \alpha_t + \beta \cdot T_{it} + \varepsilon_{it}
$$

- $\alpha_i$: unit fixed effect (person's own intercept).
- $\alpha_t$: time fixed effect (wave-specific intercept).
- $T_{it}$: 1 if person $i$ is in the treated group *and* wave $t$ is after the switch, 0 otherwise.
- $\beta$: the effect of being treated in a treated period.

. . .

This specification is called **two-way fixed effects**: one set of dummies for units, one set for time.

## What each piece absorbs

- **Unit fixed effects** $\alpha_i$: everything stable about a person (personality, baseline partisanship, deep values).
- **Time fixed effects** $\alpha_t$: everything shared across people in a given period (the state of the economy, a media event, the weather).
- **Treatment indicator** $T_{it}$: the only thing left is the "extra something" that hits treated units in treated periods.

$\beta$ captures that extra something.

## The punchline: same answer as the 2 × 2

```{r}
#| label: twfe-equal
list(
  "DiD (interaction)" = feols(
    benefit_cut_support ~ treated * post,
    data = danish_23
  ),
  "DiD + unit FE" = feols(
    benefit_cut_support ~ post + treated:post | respondent_id,
    data = danish_23
  ),
  "Two-way FE" = feols(
    benefit_cut_support ~ treated:post | respondent_id + wave,
    data = danish_23
  )
) |>
  modelsummary(
    stars = TRUE,
    coef_map = c(
      "treated:post" = "Treated x Post",
      "post:treated" = "Treated x Post",
      "post" = "Post",
      "treated" = "Treated"
    ),
    gof_map = c("nobs", "r.squared")
  )
```

All three land on the same number. In the simple 2 × 2 case, **two-way fixed effects is difference-in-differences**.

## Why bother with the two-way-fixed-effects form?

Because it **scales** to situations the interaction form can't handle easily:

- more than two waves
- more than two groups
- different units getting treated at different times (*staggered* treatment)

. . .

With staggered treatment there are subtleties we won't cover in this course — there's a recent literature (around 2020–2023) that fixes problems with the classical two-way fixed effects estimator when treatment timing varies across units. 

## Wrap-up

- Started with a **picture**: treated and control group, before and after. The widening of the gap *looked like* a party-switch effect.
- **Difference-in-differences** formalizes that intuition: the 2 × 2 table, or equivalently a regression with an interaction term.
- **Unit fixed effects** sharpen the design by isolating within-person variation: every stable personal characteristic is absorbed, so it cannot bias the treatment coefficient.
- **Time fixed effects** generalize the `post` dummy to many periods and absorb anything that hit everyone at the same moment.
- **Two-way fixed effects** = unit + time fixed effects. In the simple 2 × 2 case, it's the same number as the interaction regression, but it scales.
- **Parallel trends** remains the crucial, untestable assumption behind all of these.
- **Even without parallel trends, fixed effects are a good idea.** You make comparisons within groups that are more similar (different years within the same person, different cities within the same region, ...), so even if we can't claim causality, we can get a cleaner estimate of the association between treatment and outcome.
- **Final project.** Typically, you run a very simple regression (`outcome ~ treatment`), then add fixed effects (`outcome ~ treatment | id, region, ...`), then add controls (`outcome ~ treatment + controls | id, region, ...`) and compare results.