---
title: "Lecture 23: Spatial analysis"
format: revealjs
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false
# install.packages(c("sf", "rnaturalearth", "rnaturalearthdata", "maps", "patchwork"))
library(tidyverse)
library(sf)
library(rnaturalearth)
library(maps)
library(patchwork)
```

# Spatial data as tables

## The recipe, again

- Take unstructured data.
- Turn it into a table.
- Apply the regression and visualization tools you already know.

Today's flavor: **spatial data**.

## Why political scientists care

- Electoral districts and gerrymandering
- Conflict events (ACLED), protest locations
- Migration flows, border crossings
- "Distance to capital" or "distance to border" as regressors
- Neighborhood-level policy effects

. . .

Once the data is in a table, everything else in your toolkit applies.

# Two geometry types

## Polygons vs. points

- **Polygons** have an interior and a boundary.
  Countries, districts, census tracts, watersheds.
- **Points** are single locations.
  Events, cities, geocoded respondents.
- Lines (roads, rivers) exist too — we won't cover them today.

You all handle them with the `sf` package; basically, a `data.frame` with special map columns.

## An sf object *is* a data frame

```{r}
#| echo: true
# the "rnaturalearth" package provides map data; we'll use it
countries <- ne_countries(scale = "medium", returnclass = "sf")
countries

# modifiy the data frame as usual
countries <- countries |>
  mutate(gdp_pc = gdp_md * 1e6 / pop_est)

countries |>
  select(admin, continent, pop_est, gdp_pc) |>
  head(3)
```

The only new column is `geometry`. Every tidyverse verb works unchanged.

## Let's plot this! 

```{r}
#| echo: true
#| fig-height: 3.8
countries |>
  ggplot(aes(fill = gdp_pc)) +
  geom_sf(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    name = "GDP per capita (USD)",
    trans = "log10",
    labels = scales::comma
  )
```

## Points from lat/lon

Any data frame with coordinates becomes an sf object via `st_as_sf()`:

```{r}
#| echo: true
# likewise, the maps package provides city data; we'll use it for points
capitals <- maps::world.cities |>
  # select only capital cities
  filter(capital == 1) |>
  # convert to sf object
  st_as_sf(
    # indicate which columns hold the coordinates
    coords = c("long", "lat"),
    # indicate the coordinate reference system (CRS) of the input data
    # this data uses lat/lon in degrees, so we specify the WGS84 geographic CRS
    crs = 4326
  )

capitals |>
  select(name, country.etc, pop) |>
  head(3)
```

Same structure: rows of attributes, plus a `geometry` column.

## Let's plot the points too!

```{r}
#| echo: true
ggplot(capitals) +
  geom_sf(color = "steelblue", size = 1)
```

# The difficulties of working with spatial data

## Various file formats

- **Shapefile** (`.shp` + `.dbf` + `.shx` + ...). Old ESRI standard; multi-file; 10-char column-name limit.
- **GeoJSON** (`.geojson`). Plain text; web-friendly; one file.
- **GeoPackage** (`.gpkg`). Modern SQLite-based single-file standard. Prefer this.
- **KML / KMZ**. Legacy Google Earth.

The `sf` package can read and write all of these with `st_read()` and `st_write()`.

## Various coordinate reference systems (CRS)

The earth is not flat. Every map is a compromise between:

- preserving **angles** (shape),
- preserving **areas**,
- preserving **distances**.

A *coordinate reference system* (CRS) tells you which compromise was made.

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
p1 <- ggplot(countries) +
  geom_sf(fill = "steelblue", color = "white", linewidth = 0.1) +
  labs(title = "WGS84 (lat/lon)") +
  theme_minimal(base_size = 12)

countries_robinson <- st_transform(countries, "+proj=robin")
p2 <- ggplot(countries_robinson) +
  geom_sf(fill = "steelblue", color = "white", linewidth = 0.1) +
  labs(title = "Robinson (equal-area)") +
  theme_minimal(base_size = 12)

p1 + p2
```


## Why should we care about CRSs? 

- Geospatial data does not make sense if you don't know the CRS used to report it. 
- `sf` converts everything to a common CRS under the hood. 
- Most modern formats (GeoJSON, GeoPackage) include CRS metadata, but some other formats do not.
- If you encounter issues like points not appearing on the map, or not appearing where they should, it's likely a CRS mismatch.

# Spatial joins

## The question

> Which country does each capital fall in?

Point-in-polygon. The operation that lets point-level data inherit polygon-level attributes:

- country GDP,
- district partisanship,
- region dummies.

## `st_join()`

```{r}
#| echo: true
countries_trim <- countries |>
  select(country = admin, continent, gdp_pc)

capitals_joined <- st_join(capitals, countries_trim, join = st_within)

capitals_joined
```

## Other predicates

- `st_within` — point inside polygon
- `st_intersects` — any overlap (polygons with polygons)
- `st_contains` — polygon wholly contains another
- `st_touches` — shared boundary, no interior overlap

Pick the predicate that matches the question.

# Distance matrix

## `st_distance()`

```{r}
#| echo: true
african_caps <- capitals_joined |>
  filter(continent == "Africa") |>
  select(name, country)

D <- st_distance(african_caps, african_caps)
dim(D)
```

An $n \times n$ matrix of pairwise distances, with units attached.

## Political-science uses

- **Distance to capital** as a regressor (common in gravity and state-capacity papers).
- **Distance to border** in migration work.
- **Distance to nearest conflict event** to measure exposure or spillovers.

# Wrap-up

## What to take home

- Spatial data is a **table with a geometry column**.
- **CRS matters.** Geographic for storage, projected for computation.
- **Bins beat gradients.** Normalize counts by population.
- `sf` is your entry point.

