---
title: "Lecture 22: Text analysis"
format: revealjs
---

```{r}
#| label: setup
#| message: false
#| warning: false
# install.packages(c("tidyverse", "tidytext", "quanteda", "stm", "ggwordcloud", "httr2", "tinytable"))
library(tidyverse)
library(tidytext)
library(quanteda)
library(quanteda.textstats)
library(stm)
library(ggwordcloud)
library(httr2)
library(tinytable)

theme_set(theme_minimal(base_size = 14))
```

## Before we start

Copy the file `.Renviron.example` to `.Renviron` and add your Gemini API key.

## Our next lectures: new sources of data

- Today: text
- Network data
- Geospatial data
- Images and video (we won't do this)

The idea is always the same: turn this unstructured data into structured data (i.e., a table), then analyze the table with the tools you already know.

## 60 speeches, three ways to turn them into data

Today we treat text as data:

- A 1960s tool: **dictionaries** (count positive and negative words)
- A 2000s tool: **statistical topic models** (let the data find patterns)
- A 2020s tool: **large language models** (just ask)

Same corpus, same paragraphs. How much do the three tools agree?

## Where text shows up in political science

- Presidential speeches, party manifestos, legislative debates
- Tweets, news, campaign ads
- Court opinions and legal filings
- Open-ended survey responses

Core move: turn text into numbers, then analyze the numbers.

## The corpus: US inaugural addresses

```{r}
#| label: load-corpus
#| echo: true
inaugural <- data_corpus_inaugural
length(inaugural)
range(docvars(inaugural, "Year"))
```

```{r}
#| label: sample-quote
#| echo: true
as.character(inaugural["1961-Kennedy"]) |>
  str_wrap() |>
  str_sub(1, 250)
```

## From speeches to paragraphs

60 speeches is too few for some tools. Split each speech into paragraphs; drop very short ones.

```{r}
#| label: reshape
#| echo: true
paras <- corpus_reshape(inaugural, to = "paragraphs")
paras <- paras[ntoken(paras) >= 20]

length(paras)
```

## Preparing a word-count table

```{r}
#| label: dfm
# Words common in political oratory that drown out everything else
filler <- c(
  "will",
  "shall",
  "upon",
  "must",
  "may",
  "can",
  "us",
  "let",
  "one",
  "every",
  "now",
  "also",
  "well",
  "even",
  "ever",
  "never",
  "make",
  "made",
  "first",
  "much",
  "many",
  "still",
  "yet",
  "long",
  "whole",
  "part",
  "common",
  "just",
  "right",
  "best",
  "high",
  "without",
  "today",
  "among",
  "know",
  "way",
  "time",
  "years",
  "year",
  "day",
  "days",
  "people",
  "government",
  "nation",
  "country",
  "union",
  "states",
  "state",
  "american",
  "americans",
  "america",
  "united",
  "president",
  "fellow",
  "citizens",
  "citizen",
  "public",
  "national",
  "administration",
  "principles",
  "duty",
  "duties",
  "power",
  "support",
  "interests",
  "confidence",
  "hope",
  "faith",
  "life",
  "spirit",
  "great",
  "good",
  "new",
  "old",
  "system",
  "history",
  "future",
  "men",
  "man"
)

toks <- paras |>
  tokens(remove_punct = TRUE, remove_numbers = TRUE) |>
  tokens_tolower() |>
  tokens_remove(c(stopwords("en"), filler)) |>
  tokens_select(min_nchar = 4)

dfm_paras <- dfm(toks)
dfm_paras
```

A table of paragraphs (rows) by words (columns), filled with counts. Word order is gone — just frequencies.

# Sentiment: the dictionary approach

## The idea

A **dictionary** is a pre-made list of words tagged as positive or negative.

- Tokenize each paragraph.
- Look up each word in the dictionary.
- Count positives minus negatives.

Transparent and reproducible. Requires nothing but a word list.

## Scoring each paragraph

```{r}
#| label: sentiment-score
#| echo: true
para_df <- tibble(
  doc_id = docnames(paras),
  Year = docvars(paras, "Year"),
  President = docvars(paras, "President"),
  text = as.character(paras)
)

para_sentiment <- para_df |>
  unnest_tokens(word, text) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  count(doc_id, Year, President, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(
    total = positive + negative,
    rate = (positive - negative) / pmax(total, 1)
  )
```

## Sentiment over 230 years

```{r}
#| label: fig-sentiment
#| fig-height: 4
speech_sentiment <- para_sentiment |>
  group_by(Year, President) |>
  summarise(
    pos = sum(positive),
    neg = sum(negative),
    rate = (pos - neg) / (pos + neg),
    .groups = "drop"
  )

ggplot(speech_sentiment, aes(Year, rate)) +
  geom_vline(
    xintercept = c(1861, 1917, 1941, 2001),
    linetype = "dashed",
    color = "grey70"
  ) +
  geom_line(color = "grey60") +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue") +
  labs(
    x = NULL,
    y = "Net positive share",
    title = "Net positive word share in US inaugural addresses",
    subtitle = "Dashed lines: 1861, 1917, 1941, 2001"
  )
```

## What the dictionary misses

- **Negation**: "not good" counts as positive.
- **Sarcasm**: "great job" may be negative in context.
- **Domain meaning**: "war" is always negative, even when honoring veterans.

The dictionary is fast and honest — but it reads one word at a time.

# Let the data find the topics

## The idea

A **statistical topic model** looks for groups of words that tend to appear together across paragraphs.

- We pick a number of topics, K.
- The model learns which words cluster and which paragraphs use each cluster.
- We get back: (1) the top words for each topic, (2) each paragraph's mix of topics.

Nothing is labeled by hand. Humans interpret the clusters afterward.

## Fitting the model

```{r}
#| label: stm-fit
#| cache: true
#| message: false
stm_dfm <- dfm_paras |>
  dfm_trim(min_docfreq = 10)
stm_dfm <- stm_dfm[rowSums(stm_dfm) > 0, ]

stm_in <- convert(stm_dfm, to = "stm")

topic_model <- stm(
  documents = stm_in$documents,
  vocab = stm_in$vocab,
  K = 6,
  init.type = "Spectral",
  seed = 2026,
  verbose = FALSE
)
```

## Top words per topic, with human labels

```{r}
#| label: fig-wordclouds
#| fig-height: 5
#| fig-width: 10
topic_labels <- c(
  "1. Renewal & opportunity",
  "2. Foreign relations",
  "3. Institutions & political thought",
  "4. The presidential oath",
  "5. Freedom & democracy",
  "6. Law & economy"
)

top_terms <- tidytext::tidy(topic_model, matrix = "beta") |>
  group_by(topic) |>
  slice_max(beta, n = 20) |>
  ungroup() |>
  mutate(topic_label = factor(topic_labels[topic], levels = topic_labels))

ggplot(top_terms, aes(label = term, size = beta, color = beta)) +
  geom_text_wordcloud(rm_outside = TRUE) +
  scale_size_area(max_size = 9) +
  scale_color_viridis_c() +
  facet_wrap(~topic_label, nrow = 2) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))
```

The six labels are ours. The model only gives us the word clusters — we squint at them and decide what they mean.

## Topic prevalence over time

```{r}
#| label: fig-topics
#| fig-height: 4
topic_props <- topic_model$theta |>
  as_tibble(.name_repair = ~ paste0("topic_", seq_along(.x))) |>
  mutate(Year = docvars(stm_dfm, "Year")) |>
  pivot_longer(starts_with("topic_"), names_to = "topic", values_to = "prop") |>
  mutate(
    topic_num = as.integer(str_remove(topic, "topic_")),
    topic_label = factor(topic_labels[topic_num], levels = topic_labels)
  ) |>
  group_by(Year, topic_label) |>
  summarise(share = mean(prop), .groups = "drop")

ggplot(topic_props, aes(Year, share, fill = topic_label)) +
  geom_area(position = "fill") +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    x = NULL,
    y = "Share of paragraph content",
    fill = NULL,
    title = "What inaugural addresses talked about, by year"
  )
```

## What the topic model misses

- K is a guess. Pick 8 vs 15 and the story shifts.
- Topics are nameless — you label them by squinting at word lists.
- Results wiggle with the random seed and preprocessing choices.

Still useful: it surfaces structure you did not have to specify.

# What if we just asked a model?

## The idea

Modern language models read paragraphs in context. We can:

- Give the model a list of topics.
- Show it two or three examples.
- Ask it to classify new paragraphs and return structured output.

No training. No word counts. Just an API call.

## Authenticating

```{r}
#| label: gemini-setup
key <- Sys.getenv("GEMINI_KEY") # loaded from .Renviron — never hard-code

endpoint <- paste0(
  "https://generativelanguage.googleapis.com/v1beta/",
  "models/gemini-2.5-flash:generateContent"
)

base_req <- request(endpoint) |>
  req_headers(
    "x-goog-api-key" = key,
    "Content-Type" = "application/json"
  ) |>
  req_retry(
    max_tries = 5,
    is_transient = \(resp) resp_status(resp) %in% c(429, 500, 503)
  ) |>
  req_throttle(rate = 15 / 60)
```

## One paragraph, one request

```{r}
#| label: pick-example
obama <- paras |> corpus_subset(Year == 2009 & President == "Obama")
idx <- which(str_detect(as.character(obama), "crisis"))[1]

example_id <- docnames(obama)[idx]
example_text <- as.character(obama)[idx]
```

> Obama, 2009: `r example_text`

## Building the request body

```{r}
#| label: gemini-body
topics_enum <- c(
  "economy",
  "foreign_policy",
  "national_unity",
  "liberty_and_rights",
  "religion_and_values",
  "government_and_reform",
  "other"
)

prompt <- paste0(
  "Classify this paragraph from a US presidential inaugural address ",
  "into exactly one topic from the list.\n\n",
  'Example: "Let every nation know ... we shall pay any price" ',
  '-> {"topic": "foreign_policy", "confidence": 0.95}\n\n',
  "Paragraph: ",
  example_text
)

body <- list(
  contents = list(list(parts = list(list(text = prompt)))),
  generationConfig = list(
    responseMimeType = "application/json",
    responseSchema = list(
      type = "object",
      properties = list(
        topic = list(type = "string", enum = as.list(topics_enum)),
        confidence = list(type = "number")
      ),
      required = list("topic", "confidence")
    )
  )
)
```

## Sending the request

```{r}
#| label: gemini-call
#| cache: true
resp <- base_req |>
  req_body_json(body) |>
  req_perform()

raw <- resp_body_json(resp)
str(raw, max.level = 4)
```

The answer is buried in `raw$candidates[[1]]$content$parts[[1]]$text` — and it is itself a JSON string, so we parse it twice.

## Extracting the answer

```{r}
#| label: gemini-parse
answer_text <- raw$candidates[[1]]$content$parts[[1]]$text
answer_text

llm_answer <- jsonlite::fromJSON(answer_text)
llm_answer
```

## Three tools, one paragraph

> Obama, 2009: `r example_text`

```{r}
#| label: tbl-comparison
dict_score <- para_sentiment |>
  filter(doc_id == example_id) |>
  pull(rate)

stm_idx <- which(docnames(stm_dfm) == example_id)
stm_topic_num <- which.max(topic_model$theta[stm_idx, ])

tibble(
  Method = c("Dictionary (bing)", "Statistical topic model", "Gemini"),
  Output = c(
    sprintf("net positive share = %.2f", dict_score),
    sprintf(
      "%s (%.0f%%)",
      topic_labels[stm_topic_num],
      100 * topic_model$theta[stm_idx, stm_topic_num]
    ),
    sprintf("%s (confidence %.2f)", llm_answer$topic, llm_answer$confidence)
  )
) |>
  tt()
```

The dictionary reads word-by-word; the topic model picks up the *rhetorical register* (renewal, opportunity); the LLM names a *subject matter* (the economy).

## Trade-offs

| Method | Strengths | Weaknesses |
|---|---|---|
| Dictionary | Transparent, free, fast | Reads one word at a time |
| Statistical topic model | Finds structure unsupervised | You pick the number of topics and name them |
| LLM via API | Reads in context, flexible | Opaque, costs money, less reproducible |

No tool is best in general. Pick the one that matches your question.

## Takeaways

- Text analysis = *turn text into structured data, then use the tools you already know.*
- The three generations of tools make different assumptions; they are complements, not replacements.
- LLMs are powerful but not magic — the dictionary and the topic model are still in your kit.
